sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-10 doc/paper.Rmd
Edits

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

@@

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some of the flexibility for the particular class of models discussed here.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.
<!-- User interfaces through software such as VAST or inlabru can greatly reduce these challenges. VAST is built on TMB, and is extremely powerful (e.g., fitting many multivariate models not described here), but does not provide a simple, familiar, modular interface for users with basic spatial or spatiotemporal data. inlabru provides a somewhat more familiar interface to a large class of models including GRMF models, but is strictly Bayesian, lacks some flexibility for the particular class of models discussed here. It also relies on INLA, and at least for some problems, TMB can be much faster [@osgood-zimmerman2021]. We provide a more detailed comparison of related software packages in Table 1 and the discussion. -->

@@

<!-- -   [@anderson2019] -->
<!-- -   [@maureaud2021] -->
<!-- -   [@anderson2020] -->
<!-- -   [@osgood-zimmerman2021] -->
<!-- -   [@breivik2021] -->
<!-- -   [@monnahan2021] -->
<!-- -   [@shelton2014] -->
<!-- -   [@lindgren2011] -->
<!-- -   [@latimer2009] -->

<!-- Packages: -->

<!-- spaMM: @rousset2014 -->
<!-- spNNGP: @finley2021 -->

@@

<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).

>>>>

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, both due to measured and unmeasured variables [@cressie1993; @cressie2011].
While measured variables can be accounted for via predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
<!-- ecology has a latitude gradient: von Humboldt, A. (1808) Ansichten der Natur mit wissenschaftlichen Erlauterungen. J.G. Cotta, Tübingen, Germany. -->
Accounting for this residual correlation is important because doing so (1) allows for valid statistical inference [@legendre1989a; @dormann2007], (2) can improve predictions compared to ignoring it [e.g., @shelton2014], and (3) can be of ecological interest itself [e.g., @thorson2019d; @barnett2021].

@@

While simple on the surface, such models quickly become computationally limiting due to the need to invert large matrices in order to keep track of covariation among data [e.g., @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to Gaussian Markov random fields (GMRFs) [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can make an enormous difference to computational efficiency [@heaton2017].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as INLA and TMB, are flexible and powerful but are challenging to use for many applied ecologists.
For example, TMB requires the user to program in a C++ template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as lme4 [@bates2015] and glmmTMB [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as VAST [@thorson2019] and inlabru [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used lme4 or glmmTMB or lack some of the flexibility for the particular class of models discussed here.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

@@

@@

@@

There are many R packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2017].
sdmTMB, VAST, INLA/inlabru, and spaMM [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GMRF models; we also include spBayes as a prominent package that can fit related predictive-process models without the SPDE (Table 1).
sdmTMB, VAST, and mgcv can estimate anisotropic covariance whereas INLA/inlabru and spBayes are currently limited to isotropic covariance.
sdmTMB and mgcv focus on univariate response data, whereas VAST, INLA/inlabru, spaMM, and spBayes extend to multivariate responses with varying limitations.
To our knowledge, VAST is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016]. 
<!-- However, VAST is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
<!-- and spatial empirical orthogonal function (EOF) regression [@thorson2020]. To my reading EOF was not implemented in VAST. It seemed to me its own thing?-->
Of these packages, only sdmTMB and inlabru can currently fit threshold (e.g., hockey-stick) covariate relationships.
spaMM is limited to a single spatial random field and spBayes implements spatiotemporal fields, but only as a random walk.
<!-- TODO: still true? -->
There is considerable variability in the available observation likelihoods across packages (Table 1).
